Concurrent, Parallel Garbage Collection in Linear Time

### 2016-05-27 finaletd/mainintroduction.tex
reorganize

This chapter provides the background information regarding the problem discussed in the thesis. Section
~\ref{intro:gc} describes the fundamental idea behind necessity of garbage collection. Advanced readers can skip the section ~\ref{intro:gc}. Section ~\ref{intro:motv} discusses the advantages of the garbage collection and problem statement of the thesis. Section ~\ref{intro:contr} describes all the contributions of this thesis to solve the problem mentioned in section ~\ref{intro:motv}. Section ~\ref{intro:do} helps the readers to understand how the thesis is organized.
\section{Garbage Collection}
\label{intro:gc}
All programming languages are designed to perform complex computations. Computation contain three main ingredients : Processing unit, algorithm, and memory. Processing unit aka processor is the unit which performs arithmetic, and logical operations required. Algorithm dictates sequence of operation to performed on the data to get the desired output. Algorithm implemented in a programming language is usually referred as program and collection of programs are called software. Memory saves the input, intermediate results and final output of the computation. In this thesis, we are more concerned with object oriented programming languages and object represent user defined dynamically allocated data. Developers write algorithms in a specific programming languages. Programming language organizes the memory consumed by the algorithm. Modern programming languages uses three different memory to organize it. They are stack, static and heap memory. Static memory contains all the global variables used by the program. Stack memory is used by the programming languages to manage static(fixed size) allocation of data along with scope of the data. Heap memory is used to manage the dynamically allocated objects.

@@

Garbage collection is process of collecting all unwanted objects in the heap memory. It involves finding if all objects allocated in the heap are in use by the program.
\section{Motivation and Objectives}
\label{intro:motv}
The main advantage of the garbage collections are 
\begin{enumerate}
	\item Dangling Pointers
	\item Double free bugs
	\item Memory Leaks
\end{enumerate}
\subsection{Dangling Pointers}
	Dangling pointers are those pointers where programmer deleted the object a pointer points to. So these pointers are very dangerous in nature as one does not know what the pointer is pointing to.
\subsection{Double free bugs}
	Double free bugs is problem when an objected allocated in the heap is deleted and programmer deletes the object again. The problem with this if a new object is allocated in the space and the delete is called, the new object is deleted without knowledge. This creates some dangling pointers.
\subsection{memory leaks}
	The most important of all is that the object is allocated in the heap and it is not used. The unused object is not deleted and it consumes memory which could be reused for other objects.

@@

\subsection{Concurrent}
	Concurrent garbage collectors have negligible to zero pause time. As these type of collectors process heap along side application, application does not get locked and waiting for the garbage collector to finish processing.
\subsection{Multi-collector Garbage Collection}
	Multi-collector garbage collections are multiple independent garbage collectors that work independently on the heap to identify garabge. These collectors utilize multi-processors systems when there is lot of compute resources being under utilized.
\subsection{Global Synchronization}
	When multi-collector garbage collectors are used, conventional solutions require global barrier among all the collectors to communicate and share the information to identify garbage objects and delete them. With no global synchronization, throughput of garbage collectors will be high.
\subsection{Locality based garbage collection}
	Detecting garbage is a global predicate. Traditionally, popular methods of garbage collection algorithm involves computing garbage as global predicate. This requires scanning entire heap for better evaluating the predicate. If garbage object can be detected locally based on the a small set of object, then the garbage detection process can be parallelized.
\subsection{Scalable}
	With increase in the size of memory, conventional global garbage collection techniques will not scale with synchronization and scanning the entire heap(s). A garbage collection algorithm must be scalable to meet the future demands. Locality based garbage collection with no globally synchronized multi-collector mechanism is the way for scalability.
\subsection{Promptness}
	With global garbage collection, the cost of scanning entire heap often is very expensive. It comes with the cost that the garbage will not be collected until the scanning process starts. We require garbage collector that are prompt in collecting garbage as the heap(s) scanning might be expensive in huge distributed systems.
\subsection{Safety}
	Safety property is the most crucial part of the problem. Every garbage collector must delete only garbage object. While most garbage collector in literature satisfy this property, there are some that cannot satisfy this property. This thesis requires garbage collector to be safety in the multi-collector environment which is challenging given that there is no global synchronization.
\subsection{Completeness}
	Completeness helps garbage collector to claim that it collects all garbage objects in the heap. Reference counting garbage collector is well known for incomplete garbage colletion due to their inability to collect cyclic garbage. This property is challenging as the locality based collectors usually use reference counting. Garbage collector with localized decision making guarantee completeness is necessary to make the collector highly adaptable to enhance the performance of the applications.
\section{Contributions}
\label{intro:contr}
\section{Dissertation Organization}
\label{intro:do}

>>>>

This chapter provides the background information regarding the problem discussed in the thesis. Section
~\ref{intro:gc} describes the fundamental idea behind necessity of garbage collection. Advanced readers can skip the section ~\ref{intro:gc}. Section ~\ref{intro:motv} discusses the advantages of the garbage collection and problem statement of the thesis. Section ~\ref{intro:contr} describes all the contributions of this thesis to solve the problem mentioned in section ~\ref{intro:motv}. Section ~\ref{intro:do} helps the readers to understand how the thesis is organized.
\section{Garbage paragraph}
\label{intro:gc}
All programming languages are designed to perform complex computations. Computation contain three main ingredients : Processing unit, algorithm, and memory. Processing unit aka processor is the unit which performs arithmetic, and logical operations required. Algorithm dictates sequence of operation to performed on the data to get the desired output. Algorithm implemented in a programming language is usually referred as program and collection of programs are called software. Memory saves the input, intermediate results and final output of the computation. In this thesis, we are more concerned with object oriented programming languages and object represent user defined dynamically allocated data. Developers write algorithms in a specific programming languages. Programming language organizes the memory consumed by the algorithm. Modern programming languages uses three different memory to organize it. They are stack, static and heap memory. Static memory contains all the global variables used by the program. Stack memory is used by the programming languages to manage static(fixed size) allocation of data along with scope of the data. Heap memory is used to manage the dynamically allocated objects.

@@

Garbage collection is process of collecting all unwanted objects in the heap memory. It involves finding if all objects allocated in the heap are in use by the program.
\section{Motivation and Objectives}
\label{intro:motv}
The main advantage of the garbage collections are 
\begin{enumerate}
	\item Dangling Pointers
	\item Double free bugs
	\item Memory Leaks
\end{enumerate}
\paragraph{Dangling Pointers:}
	Dangling pointers are those pointers where programmer deleted the object a pointer points to. So these pointers are very dangerous in nature as one does not know what the pointer is pointing to.
\paragraph{Double free bugs:}
	Double free bugs is problem when an objected allocated in the heap is deleted and programmer deletes the object again. The problem with this if a new object is allocated in the space and the delete is called, the new object is deleted without knowledge. This creates some dangling pointers.
\paragraph{memory leaks:}
	The most important of all is that the object is allocated in the heap and it is not used. The unused object is not deleted and it consumes memory which could be reused for other objects.

@@

\paragraph{Concurrent:}
	Concurrent garbage collectors have negligible to zero pause time. As these type of collectors process heap along side application, application does not get locked and waiting for the garbage collector to finish processing.
\paragraph{Multi-collector Garbage :}
	Multi-collector garbage collections are multiple independent garbage collectors that work independently on the heap to identify garabge. These collectors utilize multi-processors systems when there is lot of compute resources being under utilized.
\paragraph{Global Synchronization:}
	When multi-collector garbage collectors are used, conventional solutions require global barrier among all the collectors to communicate and share the information to identify garbage objects and delete them. With no global synchronization, throughput of garbage collectors will be high.
\paragraph{Locality based garbage collection:}
	Detecting garbage is a global predicate. Traditionally, popular methods of garbage collection algorithm involves computing garbage as global predicate. This requires scanning entire heap for better evaluating the predicate. If garbage object can be detected locally based on the a small set of object, then the garbage detection process can be parallelized.
\paragraph{Scalable:}
	With increase in the size of memory, conventional global garbage collection techniques will not scale with synchronization and scanning the entire heap(s). A garbage collection algorithm must be scalable to meet the future demands. Locality based garbage collection with no globally synchronized multi-collector mechanism is the way for scalability.
\paragraph{Promptness:}
	With global garbage collection, the cost of scanning entire heap often is very expensive. It comes with the cost that the garbage will not be collected until the scanning process starts. We require garbage collector that are prompt in collecting garbage as the heap(s) scanning might be expensive in huge distributed systems.
\paragraph{Safety:}
	Safety property is the most crucial part of the problem. Every garbage collector must delete only garbage object. While most garbage collector in literature satisfy this property, there are some that cannot satisfy this property. This thesis requires garbage collector to be safety in the multi-collector environment which is challenging given that there is no global synchronization.
\paragraph{Completeness:}
	Completeness helps garbage collector to claim that it collects all garbage objects in the heap. Reference counting garbage collector is well known for incomplete garbage colletion due to their inability to collect cyclic garbage. This property is challenging as the locality based collectors usually use reference counting. Garbage collector with localized decision making guarantee completeness is necessary to make the collector highly adaptable to enhance the performance of the applications.
\section{Contributions}
\label{intro:contr}
\section{Dissertation Organization}
\label{intro:do}

### 2016-05-27 finaletd/prelim.tex
reorganize

\section{Memory Management}
Most programming languages uses stack and heap memory to organize the memory consumption of an application. Heap memory is used especially for dynamic allocated objects. During call stack unwinding, all the memory allocated in each frame is deleted there by automatically deleting unwanted memory in the frame. The idea of heap memory is also to enhance the life of the data. So a call stack unwinding does not delete all the memory allocated in the frame when it is popped. So this creates the necessity for developers to identify when the data has to be deleted and reclaim memory for reuse.

@@

\section{Abstract Graph Model}
Every block of memory that represents a data is generally referred as object in object oriented programming langauges. In this thesis, we abstract the environment into graph theoretic concepts. 
Every object allocated in the heap can be mapped in to nodes. Nodes are represent by circle in the graph. An object might contain references to other objects in the heaps. These references are unidirectional references. These references can be modeled as directed edges in the graph. So the nodes have arc followed by an arrow. The directional edge represent which objects stores the references. The arc ending with no arrow saves the references of the node with arrow. So the relationship among objects in the heap can be modeled in to nodes and edges. Roots in the stack memory and static memory can be modeled in to separate nodes or a single node. The unique property of the root node(s) (R) is/are they never contain an incoming edge. The roots are inaccesible by the objects in the heap or by other roots. So this creates a unique property among the nodes. So R only contains directional edges that points to other nodes.

\section{Properties of Garbage Collection}
\section{Types of Garbage Collection}

>>>>

\begin{comment}
\section{Memory Management}
Most programming languages uses stack and heap memory to organize the memory consumption of an application. Heap memory is used especially for dynamic allocated objects. During call stack unwinding, all the memory allocated in each frame is deleted there by automatically deleting unwanted memory in the frame. The idea of heap memory is also to enhance the life of the data. So a call stack unwinding does not delete all the memory allocated in the frame when it is popped. So this creates the necessity for developers to identify when the data has to be deleted and reclaim memory for reuse.

@@

\end{comment}

\section{Abstract Graph Model}

\begin{comment}
Every block of memory that represents a data is generally referred as object in object oriented programming langauges. In this thesis, we abstract the environment into graph theoretic concepts. 
Every object allocated in the heap can be mapped in to nodes. Nodes are represent by circle in the graph. An object might contain references to other objects in the heaps. These references are unidirectional references. These references can be modeled as directed edges in the graph. So the nodes have arc followed by an arrow. The directional edge represent which objects stores the references. The arc ending with no arrow saves the references of the node with arrow. So the relationship among objects in the heap can be modeled in to nodes and edges. Roots in the stack memory and static memory can be modeled in to separate nodes or a single node. The unique property of the root node(s) (R) is/are they never contain an incoming edge. The roots are inaccesible by the objects in the heap or by other roots. So this creates a unique property among the nodes. So R only contains directional edges that points to other nodes. 
\end{comment}

\section{Types of Garbage Collection}

\section{Cyclic Graph}
The attractive property of reference counting algorithm is localized computation and incremental approach. Mark Sweep algorithm innately requires synchronization and computing the garbage nodes globally. Although reference counting had attractive benefits, reference counting  suffered from fundamental problem of inability to detect all types of garbage especially cyclic garbage. Cyclic garbage are collection of nodes that creates a cycle. By reference counting principle, all the nodes in the graph contain non-zero reference count and hence the algorithm does not recognize the garbage. These cyclic garbage are very common. For a garbage collector to be complete, it is not enough to guarantee the safety property. Liveness property is crucial property to ensure the functional requirements of garbage collector. Brownbridge invented an idea to classify edges in the graph into two types namely : strong and weak. 
\section{Cycle detection using Strong-Weak}
Brownbridge classifies the edges in the graph into strong and weak. The classification is based on the invariant that there is no cycle of strong edges. So every cycle must contain at least one weak edge. The notion of strong edge is connected to liveness of a node. So  the existence of strong edge indicates that the node is live.

Strong edges form a connected, acyclic, directed graph in which every node is reachable from R. The rest of the edges are classified as weak. This classification is not simple to compute as themselves might requires complete scanning of the graph. To avoid complete scanning, some heuristic approaches are required. The identification of cycle creating edges are the bottleneck of the classification problem. The edges are  labeled weak if it is created to a node that already exists in the memory. this heuristic guarantee that all cycle contains at least a weak edge. But a node with weak incoming edge does not mean it is part of a cycle. The heuristic helps the mutator to save time in classifying the edges. But the collectors require more information about the topology of the subgraph to identify if the subgraph is indeed garbage to delete.
\section{Brownbridge Garbage Collector}
\section{Pitfalls of Brownbridge and successor's technique}
